---
output: html_document
---
#Assignment 1

**Course: Reproducible Research**

**Student: Nnaemeka Akpunonu**

**Due Date: Sunday, July 20, 2014**


###Task 1: Loading and preprocessing the data

**1. Load the data**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to load the data from a folder in my working directory:

```{r loadData}
actdata <- read.csv("C:/Users/nakpunonu/Google Drive/Courses Online/R Programming/R Working Directory/activitydata/activity.csv")
```


**2. Process/transform the data (if necessary) into a format suitable for your analysis**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I did not find any need to preprocess the data

###Task 2: What is mean total number of steps taken per day?

**1. Make a histogram of the total number of steps taken each day (ignore missing values)**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to draw the histogram:

```{r drawHistogram}
steps.per.day <- tapply(actdata$steps,actdata$date,sum)

hist(steps.per.day, col="blue", xlab="Total Number of Steps Per Day", main = "Histogram of Total Number of Steps Per Day",col.axis="red")
```

- As the histogram shows, the total number of steps per day for this individual falls between 0 to 25,000 steps; and mostly falls between 10,000 to 15,000 steps.

**2. Calculate and report the mean and median total number of steps taken per day.**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to compute the mean and the median:

```{r mean and median}

mean.total.steps <- mean(steps.per.day,na.rm=T)
mean.total.steps

median.total.steps <- median(steps.per.day,na.rm=T)
median.total.steps

```

- The mean total number of steps per day is `r mean.total.steps` steps
- The median total number of steps per day is `r median.total.steps` steps

###Task 3: What is the average daily activity pattern?

**1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to perform this action:

```{r time series plot}
avg.steps.per.int <- tapply(actdata$steps,actdata$interval,mean,na.rm=T)
intervals <- unique(actdata$interval)
table3 <- data.frame(interval=intervals,Avg.Steps = avg.steps.per.int)
plot(table3$interval,table3$Avg.Steps,type="l",col="blue",col.axis="red",xlab="Interval",ylab="Average Number of Steps", main = "Average Number of Steps Per Daily Interval")
```

**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to perform this action:

```{r interval with maximum steps}

table3$Avg.Steps <- as.numeric(table3$Avg.Steps)
max.step <- round(max(table3$Avg.Steps),0)
maximum <- table3$Avg.Steps==max(table3$Avg.Steps)
max.int <- table3$interval[maximum]
max.int
```

- The 5-minute interval with the maximum number of steps is interval `r max.int`. It contains an average of `r max.step` steps.

### Task 4: Imputing missing values

**1. Calculate and report the total number of missing values in the dataset (i.e. total number of rows with NA's)**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to perform this action:

```{r count of NAs}

count.of.NA <- sum(is.na(actdata$steps))
count.of.NA
```

- The total number of rows which contain NA is `r count.of.NA`.

**2. Devise a strategy to fill in all missing values in the dataset.**

- I opted to fill them in with the mean for the interval.

**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to perform this action. I also included the head and the summary of the new dataset.

```{r filling missing data}
new.actdata <- read.csv("C:/Users/nakpunonu/Google Drive/Courses Online/R Programming/R Working Directory/activitydata/activity.csv")

for(i in 1:nrow(new.actdata)){
        if(is.na(new.actdata[i,1])){
                
                new.actdata[i,1] <- table3$Avg.Steps[match(new.actdata[i,3],table3$interval)]
                
        }
}

head(new.actdata)
summary(new.actdata)
```

**4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to perform this action:

```{r histogram of new data}
new.steps.per.day <- tapply(new.actdata$steps,new.actdata$date,sum)
hist(new.steps.per.day, col="blue", xlab="Total Number of Steps Per Day", main = "Histogram of Total Number of Steps Per Day (Missing Data Imputed)",col.axis="red")
```

```{r new mean and new median}

new.mean.total.steps <- mean(new.steps.per.day,na.rm=T)
new.mean.total.steps

new.median.total.steps <- median(new.steps.per.day,na.rm=T)
new.median.total.steps
```

- The new mean total steps per day is `r new.mean.total.steps` (with missing data imputed) compared to `r mean.total.steps` without missing data imputed. 

- The new median total steps per day is `r new.median.total.steps` (with missing data imputed) compared to `r median.total.steps` without missing data imputed.

- As can be seen above, imputing missing data by using the average across respective intervals in this instance does not make any considerable difference to the analysis.

### Task 5: Are there differences in activity patterns between weekdays and weekends?

**1. Create a new factor variable in the dataset with two levels - 'weekday' and 'weekend' indicating whether a given date is a weekday or weekend day.**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to perform this action. A sample of the output dataset with the weekdays and weekday factors is shown. "1"" is used to represent weekends while "2" represents weekdays.

```{r create new factor for weekdays and weekends}

new.actdata2 <- data.frame(new.actdata)
new.actdata2$date <- as.POSIXlt(new.actdata2$date)
weekday <- weekdays(new.actdata2$date)

weekday.factor <- data.frame(weekday.factor = 1:nrow(new.actdata2))

for(i in 1:nrow(new.actdata2)){
        if(weekday[i]=="Sunday" | weekday[i]=="Saturday"){
                weekday.factor[i,1] <- 1
                } else {
                       weekday.factor[i,1] <- 2
                }      
}

weekday.factor$weekday.factor <- as.factor(weekday.factor$weekday.factor)
factor.labels <- factor(weekday.factor, levels=c(1,2), labels=c("weekends","weekdays"))

new.actdata2 <- data.frame(new.actdata2, weekday, weekday.factor)

new.actdata2[2010:2020,]

```

**2. Make a panel plot containing a time series plot of 5-minute intervals and the average number of steps taken, averaged across all weekday days or weekend days.**

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; I used the following code to perform this action.

```{r panel plots for weekdays and weekends}

factor.avg <- tapply(new.actdata2$steps,list(new.actdata2$interval,new.actdata2$weekday.factor),mean,na.rm=T,simplify=TRUE)

int <- new.actdata2[1:576,3]
factor <- factor(rep(1:2,each=288),levels=c(1,2),labels=c("weekends","weekdays"))
table5<- data.frame(steps = c(factor.avg[1:288,1],factor.avg[1:288,2]),factor=factor,interval=int)

library(lattice)

xyplot(table5$steps~table5$interval | table5$factor, data = table5, type= "l",layout=c(1,2),xlab="Intervals",ylab="Average Number of Steps",main="Average Number of Steps by Interval for Weekdays & Weekends")

```
